---
title: "LD"
created: 3/19/2019
output:
  html_document: default
  word_document: default
By: Phillip Morin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This document contains R-scripts to select genotypes based on a stratification scheme, then test the genotypes within the stratum for linkage disequilibrium (LD) and deviations from the expectations of Hardy-Weinberg Equilibrium (HWE). For each test, a p-value sequential correction for multiple tests (Holm, 1979) is applied, and the resulting adjusted p-values are output in .csv files. Other multiple-correction tests can be implemented instead of the Holm (1979) version.

A Jacknife analysis to identify genotypes responsible for deviations from HWE is added as a way to detect samples or loci that have unlikely genotypes (given HWE expectations) and may be errors. 

Holm, S. (1979), A simple sequentially rejective multiple test proceedure. Scand. J. Statist. 6:65-70. 
Morin, P.A., R.G. LeDuc, F.I. Archer, K.K. Martien, R. Huebinger, J.W. Bickham, and B.L. Taylor. 2009. Significant deviations from Hardy-Weinberg equilibirum caused by low levels of microsatellite genotyping errors. Molecular Ecology Resources 9:498-504.

# Load packages
```{r include=FALSE}
rm(list = ls())
library(strataG)
library(tidyverse)
```


# Load data files, and create gtypes object
```{r}
stratum <- "Wcoast4_num" # LD_Neah_Bay # LD_SanJuanIs
title <- "HWE_LD_Jack"
#dir.create(paste(title,"_",stratum, sep = ""))

# read genotypes data file (diploid, codominant); in directory above current directory
gdata <- read.csv("example_100_loci_genotable.csv", header = TRUE) # gdata should not have headers (missing data as NA, 00, 000, -999)
# change labid column header name
colnames(gdata)[colnames(gdata) == "id"] <- "labid"


# read coordinates file (one line per individual, 2 columns for lat, lon; no headers, no missing data)
# in directory above current directory
coord <- read.csv("Strata_example_Ppho.csv", header = TRUE, na.strings=c("","NA"))
# change labid column header name
colnames(coord)[colnames(coord) == "LABID"] <- "labid"

# combine data with coordinates (files can be different lengths)
# select the specified stratum along with labid, Latitude and Longitude from the strata file
strata.df.names <- c("labid", "Latitude", "Longitude", stratum)
coord.strat <- coord[,colnames(coord) %in% strata.df.names]
# then merge dataframes based on "labid" being a column name in both dataframes
data <- merge(coord.strat,gdata)
# remove sample data for samples not in the stratum
data2 <- data %>% drop_na(stratum)
# remove sample data if it doesn't have lat/lon
data2 <- data2 %>% drop_na(Latitude)
# Remove monomorphic loci then ane write a file of monomorphic loci for this stratification scheme.
other <- data2 %>% select(labid, Latitude, Longitude) #create dataframe to add Lat/Lon to gtypes object
strat_col <- which(colnames(data2)==stratum) # make sure correct column is used for stratum (in case Latitude and Longitude are not included in the merged data)

g <- df2gtypes(data2, #this creates a df with 2 columns for id and hap
               ploidy = 2,
               id.col = 1,
               strata.col = strat_col,
               loc.col = 5,
               other = other,
               description = title)
alleleNum<-numAlleles(g)
not.mono <- names(alleleNum)[alleleNum > 1]
not.mono.g <- g[, not.mono, ]
not.mono.df <- as.data.frame(not.mono.g)
not.mono.df$labid<-not.mono.df$ids
not.mono.df$ids<-NULL
# Add the lat/lon data back to the dataframe
data3<-merge(not.mono.df,other)
# re-order the dataframe to put Lat/Lon columns after labid's
latlon <- c("labid", "Latitude", "Longitude")
data3 <- data3[, c(latlon, setdiff(names(data3), latlon))]
data3 <- data3 %>% arrange(desc(strata), labid) # sort by species column (decreasing) so that hybrids show up on map

# save list of monomorphic loci to a .csv file
mono<-alleleNum < 2
monoloc<-names(mono) [which(mono==TRUE)]
write.csv(monoloc, paste0(stratum,"_monomorphic_loci.csv"))

#save summary of data (after excluding monomorphic loci)
data_sum <- summary(not.mono.g)
strata.df <- as.data.frame(data_sum$strata.smry)
loc.df <- as.data.frame(data_sum$locus.smry)
samp.df <- as.data.frame(data_sum$sample.smry)
write.csv(strata.df, paste0(title,"_",stratum,"_data_sum.csv"))
write.csv(loc.df, paste0(title,"_",stratum,"_loc_sum.csv"))
write.csv(samp.df, paste0(title,"_",stratum,"_sample_sum.csv"))

paste("number of individuals = ", data_sum$num.ind)
paste("number of loci = ", data_sum$num.loc)

```

# Run HWE by stratum, then calculate p-value after mutliple test correction.
(https://stat.ethz.ch/R-manual/R-devel/library/stats/html/p.adjust.html)

```{r}
hwe<-hweTest(g, use.genepop = FALSE, B = 100, show.output = FALSE,
  delete.files = TRUE, label = "HWE")
# B = the number of replicates for the Monte Carlo procedure; change to ≥1000 after testing

# add correction for multiple tests
hwe_holm_p <- p.adjust(hwe, method = "holm", n = length(hwe))
# p.adjust.methods c("holm", "hochberg", "hommel", "bonferroni", "BH", "BY",
#   "fdr", "none")

hwe_MT <- cbind(hwe, hwe_holm_p)
colnames(hwe_MT)[which(names(hwe_MT) == "hwe")] <- "hwe p-value"
write.csv(hwe_MT, paste0(title,"_",stratum,"_HWE_mt_correct.csv"))


```


# Run LD by stratum
This may take several hours to run on a personal computer. Example has reduced iterations for faster testing.
```{r}
LD <- LDgenepop(g, show.output = FALSE, delete.files = TRUE,
  label = "linkage.genepop", dem = 100, batches = 10, iter = 500)
#dem	integer giving the number of MCMC dememorisation or burnin steps. default=10000
#batches	integer giving number of MCMC batches. default=100
#iter	integer giving number of MCMC iterations. default=5000

# add correction for multiple tests
LD_holm_p <- p.adjust(LD$p.value, method = "holm", n = length(LD$p.value))
# p.adjust.methods c("holm", "hochberg", "hommel", "bonferroni", "BH", "BY",
#   "fdr", "none")
LD_MT <- cbind(LD, LD_holm_p)

write.csv(LD_MT, paste0(title,"_",stratum,"_LD_mt_correct.csv"))

```

# Jackknife to detect genotypes that affect divergence from HWE.
This may take several hours to run on a personal computer.
```{r echo=FALSE, include=FALSE, cache=TRUE}
jack.result <- jackHWE(g, exclude.num = 1, min.hwe.samples = 5, show.progress = TRUE, use.genepop = FALSE)
jack.inf <- jackInfluential(jack.result, alpha = 0.05)
if(!is.null(jack.inf$influential)) write.csv(jack.inf$influential, file = paste0(title,"_",stratum,"_influential samples.csv"), row.names = FALSE)
plot(jack.inf, main = "Cumulative frequency of odds-ratios")
# A vertical dashed line  in th eplot marks the smallest influential exclusion
```

save data
```{r}
save.image(file = paste0(title,"_",stratum,"_LD_jack.rdata"))
```

